---
title: "Assignment 2: Coding Basics"
author: "Olivia August"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r Sequence}
#1. 
sequence100 <- seq(1,100,4)   
#create a sequence starting at 1 ending at 100 with increments of 4
sequence100
#2.
mean100 <-mean(sequence100)   #determine the mean of the sequence of numbers
mean100
median100 <- median(sequence100)   #determine the median of the sequence
median100
#3. 
greatervalue <- function(mean,median){
  if(median>mean){    #check if median is larger
    "Median is larger."
  }
  else if (mean>median){   #check if mean is larger
    "Mean is larger."
  }
  else {    #if they are the same
    "Mean and median are the same."  
  }
}
greatervalue100 <- greatervalue(mean100,median100); greatervalue100 
#output whether mean or median is larger for the sequence
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r Data Frame}
# 5 & 6
VectorA<- c("Alex","Ben","Cam","David")
VectorA    #Text List
VectorB <- c(38,92,72,63)
VectorB   #Integer List
VectorC <- c("FALSE","TRUE","TRUE","TRUE")
VectorC   #Text list

#7 & 8
Class_Gradebook <- data.frame(  #create data frame
  Names = VectorA, #Name columns of the data frame
  Scores = VectorB,
  Passed = VectorC)
Class_Gradebook

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: The data frame is different from the matrix because it contains data of difference modes. For examples the names are text, but the scores are numerical.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r Passing Score}
PassingScore <- function(x){
  ifelse(x>=50,"TRUE","FAILED") #assess whether scores meet passing score of 50
  }
print(PassingScore(VectorB))  #apply function to VectorB scores
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'Ifelse' worked because the 'if' and 'else' functions only evaluated the first element of the vector.
